An investigation of factors affecting inferences from and reliability of state-space age-structured assessment models

### 2025-08-08 Project_0/manuscript/manuscript_R1.Rmd
saving work on project_0 revisions

All model selection results condition on whether all of the compared estimating models completed the optimization process without failure. We did not condition on convergence as defined by a gradient threshold or invertibity of the hessian because optimization could correctly determine an inappropriate process error assumption by estimating variance parameters at the lower bound of zero. Such an optimization could indicate poor convergence but the likelihood would be equivalent to that without the mis-specified random effects and the AIC would be appropriately higher because more (variance) parameters were estimated. All other measures of reliability described below (bias and Mohn's $\rho$) use these same criteria for inclusion of EM fits in the summarized results.

@@

@@

@@

### Spawning stock biomass and recruitment {-}

@@

@@

>>>>

All model selection results condition only on whether all of the compared estimating models completed the optimization process without failure. We did not condition on convergence as defined by a gradient threshold or invertibity of the hessian because optimization could correctly determine an inappropriate process error assumption by estimating variance parameters at the lower bound of zero. Such an optimization could indicate poor convergence but the likelihood would be equivalent to that without the mis-specified random effects and the AIC would be appropriately higher because more (variance) parameters were estimated. All other measures of reliability described below (bias and Mohn's $\rho$) use these same criteria for inclusion of EM fits in the summarized results.

@@

### Summarizing results across OMs {-}

Because of the numerous factors investigated for each of the ways we assessed reliability of the state-space assessment models, we used  two methods to summarize the most important factors that explain differences for each measure of reliability. The first method was fitting regression models with the response being each of the measures of reliability described above and predictor variables were defined based on OM and EM characteristics. For convergence results we performed logistic regressions and for AIC results we performed multinomial regressions for the type of process error that provided the lowest AIC. For other measures of reliability we fit standard linear regression models. Because relative bias and Mohn's $\rho$ for the various parameters are bounded below at -1, we used as a response the log of the absolute log-scale errors
\begin{equation}\label{bias_regression_response}
y_i = \log\left|\log\left(\widehat \theta_i\right) - \log(\left(\theta_i\right)\right|,
\end{equation}
so that lower values measure less bias whatever the sign. For relative bias we also restricted the fits to simulations where the estimated CV of the parameter or model output was less than 100% because values greater than this would typically lead the assessment scientist away from the particular configuration of the estimating model. For all regressions we fit models with factors included individually and with all factors and with all second and third order interactions. For the multinomial regression, we used the \verb|vglm| function from the VGAM package [@yee08;@yee15]. We tabulated percent reduction in residual deviance for each of regression fits.

The second method involved fitting classification and regression trees [@breimanetal84] to show how the primary factors that split the values for each measure of reliability and subsets of the measures that were alloted to each branch of the splits. Classification trees were fit to categorical measures (convergence and AIC) and regression trees were fit to the other measures with continuous scales. The response variables were the same as the regressions for the deviance reduction analyses. We used the \verb|rpart| function in the rpart package (https://cran.r-project.org/package=rpart) to fit trees and we pruned them to show only the primary branches for clarity. Generally, the factors defining primary splits for the trees corresponded to the factors with the largest reductions in deviance fitting the regression models.

@@

### Deviance reduction and classification trees {-}

### Full results {-}

@@

### Deviance reduction and regression trees {-}

We observed different effects of factors on deviance reduction for errors in estimation of the two Beverton-Holt stock-recruit parameters. When analyses were For bias of the density-dependent parameter $b$ as measured by the log absolute value of log-scale errors, deviance was not reduced by any of the individual OM or EM factors except for variation in recruitment $\sigma_R$ for R and R+S OMs. However, second and(or) third order interactions of the OM and EM factors also provided large reductions in deviance for all OM process error types which indicated that some combinations of factors could reduce bias for the $b$ parameter (Table \ref{}). Despite the explanation variation in the $b$ estimation erorrs, bias was still significant regardless of the combination of OM and EM factors (Figures regtrees and full results).

### Full results {-}

#### Spawning stock biomass {-}

@@

### Deviance reduction and regression trees {-}

### Full results {-}

@@

\begin{landscape}
\begin{figure}
\begin{center}
\includegraphics{convergence_classification_plots}
\end{center}
\caption{Classification trees indicating primary factors determining convergence as defined by providing hessian-based standard errors for R, R+S, R+M, R+Sel and R+q OMs. Lower and higher convergence rates are indicated by red and green, respectively}\label{conv_class}
\end{figure}
\end{landscape}

\begin{landscape}
\begin{figure}
\begin{center}
\includegraphics{AIC_PE_classification_plots}
\end{center}
\caption{Classification trees indicating primary factors determining which EM process error assumption provides the lowest AIC for R+S, R+M, R+Sel and R+q OMs. Lower and higher accuracy of the process error assumption are indicated by red and green, respectively}\label{AIC_PE_class}
\end{figure}
\end{landscape}

\begin{landscape}
\begin{figure}
\begin{center}
\includegraphics{SR_a_bias_regtree_plots}
\end{center}
\caption{Regression trees indicating primary factors determining reductions in sums of squares of errors in estimation measured by Eq. \ref{bias_regression_response} for the Beverton-Holt stock-recruit parameter $a$ for R+S, R+M, R+Sel and R+q OMs. Lower and higher median absolute errors of the process error assumption are indicated by green and red, respectively.}\label{SR_a_bias_regtree}
\end{figure}
\end{landscape}

\begin{landscape}
\begin{figure}
\begin{center}
\includegraphics{SR_b_bias_regtree_plots}
\end{center}
\caption{Regression trees indicating primary factors determining reductions in sums of squares of errors in estimation measured by Eq. \ref{bias_regression_response} for the Beverton-Holt stock-recruit parameter $b$ for R+S, R+M, R+Sel and R+q OMs. Lower and higher median absolute errors of the process error assumption are indicated by green and red, respectively.}\label{SR_b_bias_regtree}
\end{figure}
\end{landscape}

\begin{landscape}
\begin{figure}
\begin{center}
\includegraphics{med_M_bias_regtree_plots}
\end{center}
\caption{Regression trees indicating primary factors determining reductions in sums of squares of errors in estimation measured by Eq. \ref{bias_regression_response} for the median natural mortality rate for R+S, R+M, R+Sel and R+q OMs. Lower and higher median absolute errors of the process error assumption are indicated by green and red, respectively.}\label{med_MA_bias_regtree}
\end{figure}
\end{landscape}

\begin{table}
\caption{For each OM process error type (columns), percent reduction in deviance for logistic regression models fit to indicators of convergence with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{convergence_PRD_table}
{\input{convergence_PRD_table}}
\end{table}

\begin{table}
\caption{For each OM process error type (columns), percent reduction in deviance for multinomial logistic regression models fit to indicators of EM process error assumption with lowest AIC with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{AIC_PE_PRD_table}
{\input{AIC_PE_PRD_table}}
\end{table}

\begin{landscape}
\begin{table}
\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq. \ref{bias_regression_response} for the Beverton-Holt stock recruit relation ship parameters with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_SR_pars_PRD_table}
{\input{bias_SR_pars_PRD_table}}
\end{table}
\end{landscape}

\begin{table}
\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq. \ref{bias_regression_response} for the median natural mortality rate parameter with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_median_M_PRD_table}
{\input{bias_median_M_PRD_table}}
\end{table}

\begin{table}
\caption{For each OM process error type (columns), percent reduction in deviance for linear regression models fit to errors in estimation measured by Eq. \ref{bias_regression_response} for the terminal year spawning biomass with each OM and EM factor (rows) included individually, combined, and with second and third order interactions.}\label{bias_SSB_PRD_table}
{\input{bias_SSB_PRD_table}}
\end{table}

### 2025-08-08 Project_0/manuscript/manuscript_R1.tex
saving work on project_0 revisions

\title{An investigation of factors affecting inferences from and
reliability of state-space age-structured assessment models}
\author{Timothy J. Miller\textsuperscript{1,2} \and Greg
Britten\textsuperscript{3} \and Elizabeth N.
Brooks\textsuperscript{2} \and Gavin Fay\textsuperscript{4} \and Alex
Hansell\textsuperscript{2} \and Christopher M.
Legault\textsuperscript{2} \and Chengxue
Li\textsuperscript{2} \and Brandon Muffley\textsuperscript{5} \and Brian
C. Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
\date{02 May, 2025}

@@

All model selection results condition on whether all of the compared
estimating models completed the optimization process without failure. We
did not condition on convergence as defined by a gradient threshold or
invertibity of the hessian because optimization could correctly
determine an inappropriate process error assumption by estimating
variance parameters at the lower bound of zero. Such an optimization
could indicate poor convergence but the likelihood would be equivalent
to that without the mis-specified random effects and the AIC would be
appropriately higher because more (variance) parameters were estimated.
All other measures of reliability described below (bias and Mohn's
\(\rho\)) use these same criteria for inclusion of EM fits in the
summarized results.

@@

@@

@@

@@

@@

\hypertarget{spawning-stock-biomass-and-recruitment}{%
\subsubsection*{Spawning stock biomass and
recruitment}\label{spawning-stock-biomass-and-recruitment}}
\addcontentsline{toc}{subsubsection}{Spawning stock biomass and
recruitment}

@@

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-puntetal14}{}}%
Punt, A.E., Hurtado-Ferro, F., and Whitten, A.R. 2014. Model selection
for selectivity in fisheries stock assessments. Fisheries Research
\textbf{158}: 124--134.

@@

>>>>

\title{An investigation of factors affecting inferences from and
reliability of state-space age-structured assessment models}
\author{Timothy J. Miller\textsuperscript{1,2} \and Greg
Britten\textsuperscript{3} \and Elizabeth N.
Brooks\textsuperscript{2} \and Gavin Fay\textsuperscript{4} \and Alex
Hansell\textsuperscript{2} \and Christopher M.
Legault\textsuperscript{2} \and Chengxue
Li\textsuperscript{2} \and Brandon Muffley\textsuperscript{5} \and Brian
C. Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
\date{08 August, 2025}

@@

All model selection results condition only on whether all of the
compared estimating models completed the optimization process without
failure. We did not condition on convergence as defined by a gradient
threshold or invertibity of the hessian because optimization could
correctly determine an inappropriate process error assumption by
estimating variance parameters at the lower bound of zero. Such an
optimization could indicate poor convergence but the likelihood would be
equivalent to that without the mis-specified random effects and the AIC
would be appropriately higher because more (variance) parameters were
estimated. All other measures of reliability described below (bias and
Mohn's \(\rho\)) use these same criteria for inclusion of EM fits in the
summarized results.

@@

\hypertarget{summarizing-results-across-oms}{%
\subsubsection*{Summarizing results across
OMs}\label{summarizing-results-across-oms}}
\addcontentsline{toc}{subsubsection}{Summarizing results across OMs}

Because of the numerous factors investigated for each of the ways we
assessed reliability of the state-space assessment models, we used two
methods to summarize the most important factors that explain differences
for each measure of reliability. The first method was fitting regression
models with the response being each of the measures of reliability
described above and predictor variables were defined based on OM and EM
characteristics. For convergence results we performed logistic
regressions and for AIC results we performed multinomial regressions for
the type of process error that provided the lowest AIC. For other
measures of reliability we fit standard linear regression models.
Because relative bias and Mohn's \(\rho\) for the various parameters are
bounded below at -1, we used as a response the log of the absolute
log-scale errors \begin{equation}\label{bias_regression_response}
y_i = \log\left|\log\left(\widehat \theta_i\right) - \log(\left(\theta_i\right)\right|,
\end{equation} so that lower values measure less bias whatever the sign.
For relative bias we also restricted the fits to simulations where the
estimated CV of the parameter or model output was less than 100\%
because values greater than this would typically lead the assessment
scientist away from the particular configuration of the estimating
model. For all regressions we fit models with factors included
individually and with all factors and with all second and third order
interactions. For the multinomial regression, we used the \verb|vglm|
function from the VGAM package (Yee 2008; Yee 2015). We tabulated
percent reduction in residual deviance for each of regression fits.

The second method involved fitting classification and regression trees
(Breiman et al. 1984) to show how the primary factors that split the
values for each measure of reliability and subsets of the measures that
were alloted to each branch of the splits. Classification trees were fit
to categorical measures (convergence and AIC) and regression trees were
fit to the other measures with continuous scales. The response variables
were the same as the regressions for the deviance reduction analyses. We
used the \verb|rpart| function in the rpart package
(\url{https://cran.r-project.org/package=rpart}) to fit trees and we
pruned them to show only the primary branches for clarity. Generally,
the factors defining primary splits for the trees corresponded to the
factors with the largest reductions in deviance fitting the regression
models.

@@

\hypertarget{deviance-reduction-and-classification-trees}{%
\subsubsection*{Deviance reduction and classification
trees}\label{deviance-reduction-and-classification-trees}}
\addcontentsline{toc}{subsubsection}{Deviance reduction and
classification trees}

\hypertarget{full-results}{%
\subsubsection*{Full results}\label{full-results}}
\addcontentsline{toc}{subsubsection}{Full results}

@@

\hypertarget{deviance-reduction-and-classification-trees-1}{%
\subsubsection*{Deviance reduction and classification
trees}\label{deviance-reduction-and-classification-trees-1}}
\addcontentsline{toc}{subsubsection}{Deviance reduction and
classification trees}

\hypertarget{full-results-1}{%
\subsubsection*{Full results}\label{full-results-1}}
\addcontentsline{toc}{subsubsection}{Full results}

@@

\hypertarget{deviance-reduction-and-classification-trees-2}{%
\subsubsection*{Deviance reduction and classification
trees}\label{deviance-reduction-and-classification-trees-2}}
\addcontentsline{toc}{subsubsection}{Deviance reduction and
classification trees}

\hypertarget{full-results-2}{%
\subsubsection*{Full results}\label{full-results-2}}
\addcontentsline{toc}{subsubsection}{Full results}

@@

\hypertarget{deviance-reduction-and-regression-trees}{%
\subsubsection*{Deviance reduction and regression
trees}\label{deviance-reduction-and-regression-trees}}
\addcontentsline{toc}{subsubsection}{Deviance reduction and regression
trees}

We observed different effects of factors on deviance reduction for
errors in estimation of the two Beverton-Holt stock-recruit parameters.
When analyses were For bias of the density-dependent parameter \(b\) as
measured by the log absolute value of log-scale errors, deviance was not
reduced by any of the individual OM or EM factors except for variation
in recruitment \(\sigma_R\) for R and R+S OMs. However, second and(or)
third order interactions of the OM and EM factors also provided large
reductions in deviance for all OM process error types which indicated
that some combinations of factors could reduce bias for the \(b\)
parameter (Table \ref{}). Despite the explanation variation in the \(b\)
estimation erorrs, bias was still significant regardless of the
combination of OM and EM factors (Figures regtrees and full results).

\hypertarget{full-results-3}{%
\subsubsection*{Full results}\label{full-results-3}}
\addcontentsline{toc}{subsubsection}{Full results}

\hypertarget{spawning-stock-biomass}{%
\paragraph*{Spawning stock biomass}\label{spawning-stock-biomass}}
\addcontentsline{toc}{paragraph}{Spawning stock biomass}

@@

\hypertarget{deviance-reduction-and-regression-trees-1}{%
\subsubsection*{Deviance reduction and regression
trees}\label{deviance-reduction-and-regression-trees-1}}
\addcontentsline{toc}{subsubsection}{Deviance reduction and regression
trees}

\hypertarget{full-results-4}{%
\subsubsection*{Full results}\label{full-results-4}}
\addcontentsline{toc}{subsubsection}{Full results}

@@

\leavevmode\vadjust pre{\hypertarget{ref-breimanetal84}{}}%
Breiman, L., Friedman, J.H., Olshen, R.A., and Stone, C.J. 1984.
Classification and regression trees. Chapman; Hall/CRC, New York, NY
USA.
doi:\href{https://doi.org/10.1201/9781315139470}{10.1201/9781315139470}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-puntetal14}{}}%
Punt, A.E., Hurtado-Ferro, F., and Whitten, A.R. 2014. Model selection
for selectivity in fisheries stock assessments. Fisheries Research
\textbf{158}: 124--134.
doi:\href{https://doi.org/10.1016/j.fishres.2013.06.003}{10.1016/j.fishres.2013.06.003}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-yee08}{}}%
Yee, T.W. 2008. The {VGAM} package. R News \textbf{8}(2): 28--39.
Available from
\url{https://journal.r-project.org/articles/RN-2008-014/}.

\leavevmode\vadjust pre{\hypertarget{ref-yee15}{}}%
Yee, T.W. 2015. Vector generalized linear and additive models: With an
implementation in {R}. Springer, New York, NY USA.
doi:\href{https://doi.org/10.1007/978-1-4939-2818-7}{10.1007/978-1-4939-2818-7}.
